Rapid technological progress in white light-emitting diodes \\ and its sources in innovation and technology spillovers

### 2023-05-23 2_SSL_NE/archive/EES/article_draft_ees.tex
prepared files for nature energy submission

\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{The first commercial white light-emitting diodes (LEDs) were introduced to the market in 1996. Since then, white LEDs have experienced major improvements in performance and efficiency. Significant decreases in cost have been driven by a rapid expansion of the market share of LED-based solid-state lighting (SSL), which in turn has driven down costs even further. Today, the technology is one of the key solutions in global climate change mitigation efforts. However, despite its significance, the extent and sources of the underlying improvements have not been systematically investigated. Understanding what has driven rapid progress in LEDs can provide lessons for accelerating innovation in a broad range of other demand-side clean energy technologies. With this aim, we gather and analyse systematic evidence on cost and performance improvements in white LEDs from cost and performance modelling, augmented by a literature review and interviews with experts in industry and academia. We find that the overall efficiency of the highest performing warm white LED packages has improved from $\eta_L=5.8\%$ in 2003 to $\eta_L=38.8\%$ in 2020. During the same period, we estimate that the cost of manufacturing low-power and mid-power LED packages at a U.S. location using state-of-the-art equipment has dropped from 1.1\$ to 0.05\$ (in 2020 USD) between 2003 and 2020, a 95.5\% decrease. We further show that technology spillovers — i.e., knowledge originating in other technologies, sectors, or scientific disciplines — affected all performance dimensions of LEDs, contributing no less than 8.5\% to total efficiency improvements and $\sim$100\% to the improvements in consumer experience metrics identified, thereby playing a key role in the widespread adoption of the technology. The increase in wafer size and manufacturing yield improvements were the primary causes of LED manufacturing cost reductions.}

@@

\section{Metrics for technological progress in LEDs}

>>>>

\includegraphics{auxiliary/rsc_template_head_foot/dates} & \noindent\normalsize{§}

@@

\section{Metrics for Technological Progress in LEDs}

### 2023-05-23 2_SSL_NE/article/article_supplementary.tex
prepared files for nature energy submission

@@

Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown. Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture. In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process. Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}. The price of sapphire substrates has decreased significantly since the year 2000, as shown in the bottom panel of \cref{fig:wafers}. This can be attributed not only to the lighting industry, but more importantly to increased supply as a result of increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}. Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing. The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}. he evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.

@@

\subsection{LED Innovations}

@@

\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies, as discussed in section 5.1 and shown in Figures 4-6 in the main article. Note that phosphor-related LED innovations affecting consumer experience metrics, along with spectral efficiency, are presented separately in Table 1 in the main publication and discussed in detail in \cref{sec:innovation_phosphor}.

@@

This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics and discussed in section 5.2, Table 1 and Figures 5-7 in the main article.

@@

Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 7 in the main article). An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG). Used in combination with red-emitting sulfide phosphors, by 2002 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}. However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased. New chemically stable and non-toxic red phosphors were needed.

@@

The most recent red narrow-band phosphor innovation included in Table 1 and Figure 7 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}. The first U.S. patent application for this phosphor was filed in 2017 \cite{seibald2019phosphor}. The SALON phosphor is a derivate of the SLA phosphor. Therefore, it is the only innovation related to consumer experience metrics identified in our study that seemingly not involved technology spillovers.

@@

\paragraph{Spectral Data for Identified LED Phosphor Innovations}

\begin{figure}[h!]
	\centering
    \includegraphics[width=0.95\textwidth]{2_SSL_EST/article/figures/phosphor_spectrum-comparison.pdf}
	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 7 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year. Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 7 in the main article. The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow. Metrics were calculated from spectral data shown in the two bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}. Bottom two rows of panels: Corresponding spectral data. The luminosity function \cite{cie-term-lumeff}, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel. Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel. Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: ? -  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, G$^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, O$^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}. Source (top panel): own elaboration based on spectral data. Arb. Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum. Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
\label{fig:phosphor_spectrum}
\end{figure}

\clearpage
\subsection{Cost Comparison with Reported Industry Data and DOE Projections}

@@

\begin{figure}[h]
	\centering
    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_calibration.pdf}
	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel). Hatched bars are projections given by US DOE for 2018 and 2020. Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}. Sources for bottom panel: own elaboration based on the cost model described in Section \cref{sec:costmodel}.}
	\label{fig:costmodel_calibration}
\end{figure}

@@

\subsection{Figure 4 (Historical Developments in Device Sub-Efficiencies)}

The sources for Figure 4 in the main article are grouped by sub-efficiencies represented on different panels of the figure:

>>>>

\setcounter{secnumdepth}{4}

@@

Sapphire wafers form the substrate on which all other layers of the light-emitting diode are grown. Being transparent to radiation in the visible spectrum, it is not removed after growth in the Classical architecture or the Thin-Film Flip-Chip architecture. In the Vertical Thin-Film architecture, it is removed by means of a laser-lift-off process. Wafers can be either unpatterned or patterned, where the latter has become commonplace by 2020 due to the beneficial properties that microstructures on the surface have on layer growth \cite{wuu2009defect} and light-extraction efficiency \cite{lee2006enhancing}. The price of sapphire substrates has decreased significantly since the year 2000, as shown in the bottom panel of \cref{fig:wafers}. This can be attributed not only to the lighting industry, but more importantly to increased supply as a result of increased demand from electronics manufacturing \cite{yole2015sapphire}, where sapphire glass is used to protect screens and sensor interfaces from scratches \cite{khattak2016world}. Wafer sizes used in manufacturing have also increased, due to the favourable economics of large wafer processing. The market has been dominated by U.S.-based \textit{Rubicon Technology} and Russian-based \textit{Monocrystal}. The evolution of sapphire wafer prices and diameters used in LED manufacturing over the years is shown in top panel of \cref{fig:wafers}.

@@

\clearpage
\subsection{LED Innovations}

@@

\cref{tab:innovations} summarizes LED innovations and technology and manufacturing process improvements identified in this study as affecting key white LED sub-efficiencies. Note that phosphor-related LED innovations affecting consumer experience metrics, along with spectral efficiency, are presented separately in Table 1 in the main publication and discussed in detail in \cref{sec:innovation_phosphor}.

@@

This section provides additional details for phosphor-related LED innovations identified in this study as affecting both spectral sub-efficiencies and consumer experience metrics.

@@

Importantly, the YAG phosphor does not exhibit the spectral properties desirable in general illumination applications (see Figure 6 in the main article). An early solution to this problem, which was first discovered in the late 1967 \cite{holloway1969optical} and suggested for LEDs by the same team at Nichia in 1996 \cite{bando1998development}\cite{shimizu1999light}, was to use the gadolinium-doped red-shifted YAG phosphor (YGAG). Used in combination with red-emitting sulfide phosphors, by 2002 it helped bring to the market the first generation of warm white light LED products, e.g., those produced by Lumileds \cite{Mueller2002}. However, sulfide phosphors led to accelerated deterioration of sensitive parts of LED devices and became less efficient as operating temperatures increased. New chemically stable and non-toxic red phosphors were needed.

@@

The most recent red narrow-band phosphor innovation included in Table 1 and Figure 6 in the main article, indicated as SALON, has been under development during the late 2010s by a group of Austrian and German researchers that included Huppertz, the discoverer of the “258” material, working in collaboration with Osram, another major LED manufacturer \cite{seibald2019phosphor}\cite{Hoerder2019}\cite{Hoerder2020}. The first U.S. patent application for this phosphor was filed in 2017 \cite{seibald2019phosphor}. The SALON phosphor is a derivate of the SLA phosphor. Therefore, it is the only innovation related to consumer experience metrics identified in our study that seemingly not involved technology spillovers.

@@

\clearpage
\paragraph{Spectral Data for Identified LED Phosphor Innovations}

\hspace{1mm}

\begin{figure}[H]
	\centering
    \includegraphics[width=0.9\textwidth]{2_SSL_EST/article/figures/phosphor_spectrum-comparison.pdf}
	\caption{Spectral data and additional consumer experience metrics for the earliest identified representative white LED products with published spectral data, shown in Figure 6 in the main article, that used phosphor innovations listed in Table 1 in the main article, each indicated by the phosphor label and spectral data publication year. Top panel: Luminous efficacy of radiation (LER) and colour rendering index (CRI) of white LED devices represented in Figure 6 in the main article. The desirable direction of improvements towards higher luminous efficacy at higher CRI is indicated by a red arrow. Metrics were calculated from spectral data shown in the two bottom panels using the \texttt{colour-science} package for Python \cite{colour-science_software}. Bottom two rows of panels: Corresponding spectral data. The luminosity function \cite{cie-term-lumeff}, describing the wavelength-dependent sensitivity of the human eye, is shown for reference in red in each panel. Note that peaks or large tails of the device emission spectrum at the far ends of the luminosity function are not desirable, as the photons of the corresponding energy are lost to the human eye and count towards the spectral loss channel. Plot legends indicate the years of publication of the spectral data and phosphor mixtures used in corresponding LED devices, with the following designations for additional phosphor mix components: ? -  other parts of phosphor mixture not disclosed, G$^*$ - \ch{CaSrS:Eu^{2+}}, Y$^\star$ - $\beta-$SiAlON:Eu$^{2+}$, G$^\dagger$ - \ch{Lu_3Al_5O_{12}:Ce^{3+}}, O$^\ddagger$ - \ch{(Ba,Sr)_2Si_5N_8:Eu^{2+}}. Source (top panel): own elaboration based on spectral data. Arb. Units - Arbitrary Units, defined as the ratio of radiation intensity at a wavelength compared to the highest point in the spectrum. Sources: top panel - own elaboration based on spectral data; bottom two panels: adapted from published spectral data for LEDs with the following phosphors: YAG \cite{bando1998development}, YGAG \cite{Mueller2002}, 258 \cite{MuellerMach2005}, SLA \cite{Pust2014}, PFS \cite{trigain_spectrum}, QD \cite{lumileds2016qd}\cite{osram2019qd}, SALON \cite{Hoerder2019}.}
\label{fig:phosphor_spectrum}
\end{figure}

\newpage
\subsection{Cost Comparison with Reported Industry Data and DOE Projections}

@@

\begin{figure}[h]
	\centering
    \includegraphics[width=\textwidth]{2_SSL_EST/article/figures/costmodel_calibration.pdf}
	\caption{Comparison of LED manufacturing cost structure between our manufacturing cost model (bottom panel) and previously published US DOE manufacturing cost calculations and projections based on industry data and LEDCOM model (top panel). Hatched bars are projections given by US DOE for 2018 and 2020. Sources for top panel: \cite{doe2010solid}\cite{doe2011solid}\cite{doe2012solid}\cite{doe2013solid}\cite{doe2014solid}\cite{doe2015solid}\cite{doe2016solid}. Sources for bottom panel: own elaboration based on the cost model described in \cref{sec:costmodel}.}
	\label{fig:costmodel_calibration}
\end{figure}

@@

\subsection{\cref{fgr:subeff} (Historical Developments in Device Sub-Efficiencies)}

The sources for \cref{fgr:subeff} are grouped by sub-efficiencies represented on different panels of the figure:
